---
title: "Life expectancy"
author: "Yufei Liu"
thanks: "Code and data are available at: https://github.com/Florence-Liu/life_expectancy"
date: today
date-format: long
abstract: ""
format: pdf
number-sections: true
toc: false
bibliography: references.bib
---

# Introduction




# Data {#sec-data}

We used `R` to do the analysis in this paper [@citeR]. We used packages `tidyverse`[@citeTidyverse], `janitor`[@citeJanitor], and `here`[@citeHere] to clean and load the data as well as create figures, and `knitr`[@citeKnitr] and `modelsummary`[@citeModelsummary] to generate tables. The color style of the figures was created referring to a R colors cheet-sheet [@citeRcolor].

## Data description 

The datasets used in this paper was obtained from World Health Organization (WHO) Global Health Observatory data repository, and is publicly available from WHO website [@citeWHO]. We utilized 3 datasets: life expectancy and Health life expectancy by WHO region[@citeLife], SDG Target 3.5 Substance abuse by WHO region [@citeAlcohol], and SDG Target 3.a Tobacco control by WHO region [@citeTobacco]. The life expectancy dataset contains life expectancy and helthy life expectancy at birth and at age 60 for different sexes and WHO regions in year 2000, 2010, 2015, and 2019. 
The substance abuse data contains 



## Data Visualization

```{r}
#| echo: false
#| message: false
#| warning: false


#### Workspace setup ####
library(tidyverse)
library(knitr)
library(here)
library(modelsummary)

#### Read in the cleaned data ####
data <- read_csv(here("outputs/data/merged_data.csv"), 
                 show_col_types = FALSE)
```

```{r}
#| label: fig-tobacco
#| fig-cap: The effect of prevalence of tobacco on life expectancy at age 60 for different sex
#| echo: false
#| message: false
#| warning: false
data |> ggplot(aes(x=prevalence_of_tobacco, y=life_expectancy, color=sex)) +
  geom_point() +
  theme_minimal() +
  labs(x = "Prevalence of tobacco use in percentage",
       y = "Life expectancy at age 60 in years",
       color = "Sex") +
  scale_color_manual(values = c("indianred3","steelblue3"))
```
```{r}
#| label: fig-alcohol
#| fig-cap: The effect of total alcohol per capita consumption on life expectanct at age 60 for different sex
#| echo: false
#| message: false
#| warning: false
data |> ggplot(aes(x=alcohol_consumption, y=life_expectancy, color=sex)) +
  geom_point() +
  facet_grid(.~sex) +
  theme_light() +
  labs(x = "Total alcohol per capita consumption in litre",
       y = "Life expectancy at age 60 in years",
       color = "Sex") +
  scale_color_manual(values = c("indianred3","steelblue3"))
```
```{r}
#| label: fig-year
#| fig-cap: Life expectancy at age 60 across years for different sex
#| echo: false
#| message: false
#| warning: false
data |> ggplot(aes(x=as.factor(year), y=life_expectancy, color=sex)) +
  geom_boxplot() +
  theme_minimal() +
  labs(x = "Year",
       y = "Life expectancy at age 60 in years",
       color = "Sex") +
  scale_color_manual(values = c("indianred3","steelblue3"))
```

```{r}
#| label: tbl-year
#| tbl-cap: Summary of average life expectancy, prevalence of tobacco, and alcohol consumption across years
#| echo: false
#| message: false
#| warning: false
table1 <- data |> group_by(year) |>
  summarise(mean_life = round(mean(life_expectancy),1), 
            mean_tobacco = round(mean(prevalence_of_tobacco),1),
            mean_alcohol = round(mean(alcohol_consumption),1))

kable(table1, col.names = c("Year","Mean<br>life expectancy (year)", 
                            "Mean<br>prevalence of tobacco (%)",
                            "Mean<br>alcohol consumption (litre)"),
      escape = FALSE, "html")
```

```{r}
#| label: tbl-region
#| tbl-cap: Summary of average life expectancy, prevalence of tobacco, and alcohol consumption for different WHO regions
#| echo: false
#| message: false
#| warning: false
table2 <- data |> group_by(region) |>
  summarise(mean_life = round(mean(life_expectancy),1), 
            mean_tobacco = round(mean(prevalence_of_tobacco),1),
            mean_alcohol = round(mean(alcohol_consumption),1))

kable(table2, col.names = c("Region","Mean<br>life expectancy (year)", 
                            "Mean<br>prevalence of tobacco (%)",
                            "Mean<br>alcohol consumption (litre)"), 
      escape = FALSE, "html")
```


# Model


# Result
```{r}
#| label: tbl-model
#| tbl-cap: Summary of two linear regression models
#| echo: false
#| message: false
#| warning: false

model1 <- lm(life_expectancy ~ prevalence_of_tobacco, data = data)
#summary(model1)


model2 <- lm(life_expectancy ~ alcohol_consumption * sex, data=data)
#summary(model2)

#plot(model)


modelsummary(list("Tobacco" = model1, 
                  "Alcohol*Sex" = model2))

```

```{r}
#| label: fig-model1
#| fig-cap: Prevalence of tobacco use versus life expectancy at age 60 with fitted regression line
#| echo: false
#| message: false
#| warning: false

data |> ggplot(aes(y=life_expectancy, x=prevalence_of_tobacco)) +
  geom_point(color="goldenrod") +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  theme_minimal() +
  labs(x = "Prevalence of tobacco use in percentage",
       y = "Life expectancy at age 60 in years") 
```

```{r}
#| label: fig-model2
#| fig-cap: Alcohol consumption per capita versus life expectancy with fitted regression line for different sex group
#| echo: false
#| message: false
#| warning: false

data |> ggplot(aes(y=life_expectancy, x=alcohol_consumption, color=sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  theme_minimal() +
  labs(x = "Total alcohol per capita consumption in litre",
       y = "Life expectancy at age 60 in years",
       color = "Sex") +
  scale_color_manual(values = c("indianred3","steelblue3"))
```


# Discussion


\newpage
# Reference

\newpage
# Appendix
```{r}
#| echo: false
#| warning: false
#| message: false

plot(model1)
plot(model2)
```